Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-01 paper.tex
Some single-cell refs and narrative
Closes #67

Although Zarr is optimised for cloud settings, it also
works very well in the classical (shared) file-system setting
and we therefore focus on benchmarks in this setting
to allow us to compare with existing tools.
Those seeking benchmarks of Zarr in cloud deployments are 
directed to [citations].

@@

Reel off Python data science ecosystem: Pandas~\citep{mckinney2010data} etc
[ Things to cite: pandas~\citep{mckinney2010data},
BioNumpy~\citep{rand2022bionumpy}, OME-Zarr~\citep{moore2023ome}]

@@

Scientific datasets are increasingly overwhelming the classical
model of downloading and analysing locally, and are migrating to 
centralised cloud repositories \citep{abernathey2021cloud}.
The combination of Zarr's simple and cloud-friendly storage 
of data chunks with state-of-the-art compression methods has 
led to Zarr gaining significant traction in these settings.
Multiple petabyte-scale datasets are now stored using 
Zarr~\cite[e.g.][]{gowan2022using, % weather model data
fahnestock2023mappin, % ITS_LIVE dataset
cmip6_dataset}
or under active consideration for migration~\citep{abernathey2021opening}.
The Open GeoSpatial consortium has formally recognised Zarr as a community
standard~\cite{ogc_zarr2_standard}
and has formed
a new GeoZarr Standards Working Group to establish a Zarr encoding for
geospatial data~\cite{ogc_geozarr_news}.

Zarr has recently been gaining traction in biological applications
also. EUGENe~\citep{klie2023predictive} is a toolkit for 
for applying deep-learning methods to genomic sequence data,
integrating with PyTorch [citation?].

Spatially resolved omics data~\citep{baker2023emobject,marconato2024spatialdata}
building on the Zarr-based OME-NGFF format~\citep{moore2023ome,rzepka2023toward}

@@

>>>>

Although Zarr is optimised for cloud settings, it also
works very well in the classical (shared) file-system setting
and we therefore focus on benchmarks in this setting
to allow us to compare with existing tools.
Those seeking benchmarks of Zarr in cloud deployments are 
directed to~\citep{durbin2020task,moore2021ome,gowan2022using} [ANY OTHERS?]

@@

The Python data science ecosystem is increasingly popular, 
with packages such as NumPy~\citep{harris2020array},
SciPy~\citep{virtanen2020scipy},
Pandas~\citep{mckinney2010data},
Jupyter~\citep{kluyver2016jupyter}
enabling much of contemporary science.

BioNumpy~\citep{rand2022bionumpy},

@@

Scientific datasets are increasingly overwhelming the classical
model of downloading and analysing locally, and are migrating to 
centralised cloud repositories \citep{abernathey2021cloud,moore2021ome}.
The combination of Zarr's simple and cloud-friendly storage 
of data chunks with state-of-the-art compression methods has 
led to Zarr gaining significant traction in these settings.
Multiple petabyte-scale datasets are now stored using 
Zarr~\cite[e.g.][]{gowan2022using, % weather model data
fahnestock2023mappin, % ITS_LIVE dataset
cmip6_dataset}
or under active consideration for migration~\citep{durbin2020task,abernathey2021opening}.
The Open GeoSpatial consortium has formally recognised Zarr as a community
standard~\cite{ogc_zarr2_standard}
and has formed
a new GeoZarr Standards Working Group to establish a Zarr encoding for
geospatial data~\cite{ogc_geozarr_news}.

Zarr has recently been gaining popularity in biological applications.
The Open Microscopy Environment has developed OME-Zarr~\cite{moore2023ome} as one 
of its ``next generation'' cloud ready file formats~\citep{moore2021ome}.
OME-Zarr already has a rich suite of supporting
tools~\cite{moore2023ome,rzepka2023toward}. 
Zarr has also seen recent uptake 
in single-scell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
and multimodal spatial omics
data~\citep{marconato2024spatialdata,baker2023emobject}.

EUGENe~\citep{klie2023predictive} is a toolkit for 
for applying deep-learning methods to genomic sequence data,
integrating with PyTorch [citation?].

@@

Zarr was originally developed as a means efficiently accessing VCF data,
 and has been in use within the MalariaGEN project for XX years. [
ALISTAIR - details and citations here!]
